\vspace{1.0cm} \Large \bf Supplemental material of ``Health and disease imprinted in the time variability of the human microbiome''

### 2017-02-21 mSystems/rev2/mSys_MandM_rev2.tex
mSys LaTeX files updated to ACCEPTED version (and beyond...)
"Beyond" means that formulae have been improved, typos corrected
and LaTeX magic comments added.

On branch master, changes committed:
	new file:   rev2/mSys_MAIN_rev2.dvi
	modified:   rev2/mSys_MAIN_rev2.pdf
	modified:   rev2/mSys_MAIN_rev2.tex
	modified:   rev2/mSys_MandM_rev2.tex
	modified:   rev2/mSys_discussion_rev2.tex
	modified:   rev2/mSys_intro_rev2.tex
	modified:   rev2/mSys_results_rev2.tex

%%
%TC:ignore
%\small
\subsection*{Model} \label{sec:model}

We modeled microbial abundances over time along the lines of Blumm \textit{et al.}\cite{ranking}. The dynamics of taxon relative abundances was described by the Langevin equation:
\begin{linenomath}
\begin{equation*}
\dot{x_i} = F_i \cdot x_i^\alpha + V \cdot x_i^\beta \xi_i(t) - \phi(t) \cdot x_i,
\end{equation*}
\end{linenomath}
where $F_i$ captured the fitness of the taxon $i$, $V$ corresponded to the noise amplitude and $\xi_i$(t) was a Gaussian random noise with zero mean  $<\xi_i(t)> = 0$, and variance which was uncorrelated over time, $<\xi_i(t) \xi_i(t')> = \delta(t'-t)$. The function $\phi(t)$ ensured the normalization at all times, $\sum x_i(t) = 1$, and corresponded to $\phi(t) = \sum F_i x_i^\alpha + \sum V x_i^\beta \xi_i(t)$.
The temporal evolution of the probability that taxon $i$ had a relative abundance $x_i(t)$, $P(x_i,t)$, was determined by the Fokker-Planck equation:
\begin{linenomath}
\begin{equation*}
\frac{\partial P}{\partial t} = - \frac{\partial}{\partial x_i}  [(F_i \cdot x_i^\alpha - \phi(t) \cdot x_i ) \cdot P]+ \frac{1}{2} \frac{\partial^2}{\partial x_i^2} (V^2 \cdot x_i^{2\beta}\cdot P).
\end{equation*}
\end{linenomath}
The microbiota evolved towards a steady state with a time-independent probability dependent on the values of $\alpha$, $\beta$, $F_i$ and $V$. For $\alpha<1$ (otherwise, systems are always unstable), the steady state probability was localized in a region around a preferred value or broadly distributed over a wide range, depending on whether the fitness $F_i$ dominated or was overwhelmed by the noise amplitude $V$. The steady-state solution of the Fokker--Planck equation was given by:
\begin{eqnarray*}
P_0 (x_i) &=& C_{ne}(\alpha,\beta,F_i,V)  \cdot x_i^{-2\beta}  \cdot \exp\Big[\frac{2F_i}{V^2}\frac{x_i^{1+\alpha-2\beta}}{1+\alpha-2\beta}-\frac{\phi_0}{V^2}\frac{x_i^{2-2\beta}}{1-\beta}\Big] \quad \textrm{if} \quad  2\beta \ne 1+\alpha, \\
P_0 (x_i) &=& C_e(\alpha,\beta,F_i,V)  \cdot x_i^{\frac{2F_i}{V^2} -2\beta}  \cdot \exp\Big[\frac{\phi_0}{V^2}\frac{x_i^{2-2\beta}}{1-\beta}\Big] \quad \textrm{if} \quad  2\beta = 1+\alpha,
\end{eqnarray*}
where $\phi_0 = (\sum_i F_i^{1/(1-\alpha)})^{1-\alpha}$ and $C_{ne}$ and $C_{e}$ were integrals that were solved numerically for the parameters of interest. The ordered phase occurred when the solution had a maximum in the physical interval ($0<x_i<1$). For a larger V, the transition to a disordered phase happened when the maximum shifted to the unphysical region $x_i<0$, which sets the phase transition region $V(\alpha,\beta,F_i)$. The phase transition region was calculated analytically in specific cases:
\begin{eqnarray*}
F_i^2 &=& 4 \beta \phi_0 V^2\quad \textrm{if} \quad  \beta = \alpha \neq 1,\\
F_i &=& \beta V^2\quad \textrm{if} \quad  2\beta = 1+\alpha,
\notag
\end{eqnarray*}
where the first case, simplified to $F = 3 V^2$ if $\beta = 0.75$ and the fitness of this taxon dominated in $\phi_0$. 
In many physical systems (Brownian motion is the classic example \cite{Einstein}), the two terms of the Langevin's equation are related.  The \emph{Fluctuation Dissipation Theorem} states out a general relationship between the response to an external disturbance and the internal fluctuations of the system\cite{FD}. The theorem can be used as the basic formula to derive the fitness from the analysis of fluctuations 
of the microbiota, assuming that it is in equilibrium (the ordered phase).

%%
\subsection*{Standardization} \label{sec:stan}
In order to properly show all the studies under common axes, we decided to standardize the Taylor parameters using the group of healthy individuals for every single study independently. With this approach, all the studies can be visualized in a shared plot with units of Taylor--parameter standard--deviation on their axes.

For a Taylor's parameter, e.g. $V$, the estimate of the mean ($\widehat{V}$) for the healthy subpopulation, composed of $h$ individuals, is:
\begin{linenomath}
$$\widehat{V} = \frac{1}{W_1}\sum_{i=1}^h V_i \omega_i=\sum_{i=1}^h V_i \omega_i$$
\end{linenomath}
as $W_1=\sum_i^h \omega_i=1$, since $\omega_i$ are normalized weights calculated as:
\begin{linenomath}
$$\omega_i = \frac{\frac{1}{\sigma^2_{V_i}}}{\sum_i^h\frac{1}{\sigma^2_{V_i}}}$$
\end{linenomath}
$\sigma_{V_i}$ being the estimation of the uncertainty in $V_i$ obtained together with $V_i$ from the X-weighted power-law fit described in Section \ref{sec:X-w}, for healthy individuals.

@@

%%
\subsubsection*{Sample selection}
We chose studies about relevant pathologies containing metagenomic sequencing time data series of bacterial populations from humans in different healthy and non-healthy states. Only those individuals who had three or more time points of data available in databases were selected. The study by Caporaso \textit{et al.} study \cite{moving} was selected as it featured two healthy individuals measured over a very long timespan, with almost daily sampling. The study of Faith \textit{et al.} \cite{LEA} was selected given the BMI differences between subjects. Moreover, some of them followed diets which could be treated as system perturbations. Only those individuals who had normal or overweight BMI were considered as healthy. The study by Smith \textit{et al.} \cite{kwashiorkor} was selected for both the age of the patients and the rare disease. We only worked with the discordant twins, and considered those who were not affected by kwashiorkor in each pair of patients as being healthy. The study by David \textit{et al.} \cite{diet} was selected for its differential diets. The healthy part was considered to be the first time samples of each individual before the diet, while the rest of the time points were considered as perturbations. Dethlefsen and Relman's work \cite{antibiotic} was selected due to the interesting treatment of two antibiotic intakes by three different subjects of the same antibiotic. The healthy part was considered to be only those times before any antibiotic treatment, and the time of antibiotic intakes and the period after that as perturbations. The work by David \textit{et al.} \cite{hostlife} was selected due to the comprehensive longitudinal data that it provides plus its complete metadata and the interesting events that happened to both subjects (an infection and a trip abroad). The healthy part was taken from time points before or after each event. Finally, we also considered a study from our group carried out by Durban \textit{et al.} \cite{IBS} in which the healthy subjects were considered as those who did not suffer from irritable bowel syndrome, while the patients who had this disease were taken as perturbations.

@@

%%
\subsubsection*{16rRNA sequences processing}
Reads from the selected studies were first quality filtered using the FastX toolkit\cite{FASTX}, allowing only those reads which had more than a quality score of over 25 in 75\% of the complete sequence. 16S rRNA reads were then clustered at a 97\% nucleotide sequence identity (97\% ID) into operational taxonomic units (OTUs), using the QIIME software package\cite{QIIME} (version 1.8). We followed an open reference OTU picking workflow in all cases. The clustering method used was UCLUST, and the OTUs were matched against the Silva database\cite{SILVA} (version 111, July 2012) and were assigned to a taxonomy with a UCLUST-based consensus taxonomy assigner. The parameters used in this step were: similarity 0.97, prefilter percent id 0.6, maximum accepts 20, maximum rejects 500.

%%
\subsubsection*{Metagenomic sequences processing}
Shotgun metagenomic sequences were analyzed with LMAT (Livermore Metagenomics Analysis Toolkit) software package\cite{LMAT} (version 1.2.4, with Feb'15 release of the LMAT-Grand database). LMAT was run using a Bull shared-memory node belonging to the team?s HPC (high performance computing) cluster. It was equipped with 32 cores (64 threads available using Intel Hyper-Threading Technology) as it has two Haswell-based Xeons (22 nm technology), the E5-2698v3@2.3 GHz, sharing half a tebibyte of DRAM memory. This node is also provided with a PCIe SSD card as NVRAM, the Micron P420m HHHL, with 1.4 TB, and 750000 reading IOPS, 4 KB, achieving 3.3 GB/s. The computing node was supplied with a RAID-0 (striping) scratch disk area. We used the ``Grand'' database\cite{LMAT2}, released in Feb'15, provided by the LMAT team, where ``Grand'' refers to a huge database that contains k-mers from all the viral, prokaryote, fungal and protist genomes present in the NCBI database, plus the Human reference genome (hg19), plus GenBank Human, plus the 1000 Human Genomes Project (HGP) (this represents about 31.75 billion k-mers occupying 457.62 GB)\cite{LMAT2}. Before any calculations were made, the entire database was loaded into the NVRAM. With this configuration, the observed LMAT sustained sequence classification rate was $20\unclassrate$. Finally, it is worth mentioning that a complete set of Python scripts  was developed as back-end and front-end of the LMAT pipeline in order to manage the added complexity of time series analysis (\href{https://github.com/DLSteam/MAUS_scripts}{[https://github.com/DLSteam/MAUS\_scripts]}).

@@

When fitting the power-law of std vs. mean, we took into account that every mean has uncertainty and can be estimated for a sample size $n$ by the SEM (\emph{Standard Error of the Mean}). Here, the uncertainties affected the independent variable, so the fit was not as trivial as a Y-weighted fit, where the uncertainties affect the dependent variable. A standard approach for this fit is: a) to invert the variables before applying the weights, b) then perform the weighted fit, and finally, c) revert the inversion. This method is deterministic, but the approximate solution worsens with smaller coefficients of determination. To overcome this limitation, we developed a stochastic method by using a bootstrapping-like strategy that avoided inversion and could be applied regardless of the coefficient of determination.

>>>>

% !TeX root = ./mSys_MAIN_rev2.tex
%TC:ignore
%\small
\subsection*{Model} \label{sec:model}

We modeled microbial abundances over time along the lines of Blumm \textit{et al.}\cite{ranking}. The dynamics of taxon relative abundances was described by the Langevin equation:
\begin{linenomath}
\begin{equation*}
\dot{x_i} = F_i \cdot x_i^\alpha + V \cdot x_i^\beta \xi_i(t) - \phi(t) \cdot x_i,
\end{equation*}
\end{linenomath}
where $F_i$ captured the fitness of the taxon $i$, $V$ corresponded to the noise amplitude and $\xi_i$(t) was a Gaussian random noise with zero mean  $\left<\xi_i(t)\right> = 0$, and variance which was uncorrelated over time, $\left<\xi_i(t) \xi_i(t')\right> = \delta(t'-t)$. The function $\phi(t)$ ensured the normalization at all times, $\sum x_i(t) = 1$, and corresponded to $\phi(t) = \sum F_i x_i^\alpha + \sum V x_i^\beta \xi_i(t)$.
The time--course changes in the probability that taxon $i$ had a relative abundance $x_i(t)$, $P(x_i,t)$, was determined by the Fokker--Planck equation:
\begin{linenomath}
\begin{equation*}
\frac{\partial P}{\partial t} = - \frac{\partial}{\partial x_i}  \left[(F_i \; x_i^\alpha - \phi(t) \; x_i ) \; P\right]+ \frac{1}{2} \frac{\partial^2}{\partial x_i^2} \left(V^2 \; x_i^{2\beta}\; P\right).
\end{equation*}
\end{linenomath}
The microbiota evolved towards a steady state with a time--independent probability dependent on the values of $\alpha$, $\beta$, $F_i$ and $V$. For $\alpha<1$ (otherwise, systems are always unstable), the steady--state probability was localized in a region around a preferred value or broadly distributed over a wide range, depending on whether the fitness $F_i$ dominated or was overwhelmed by the noise amplitude $V$. The steady--state solution of the Fokker--Planck equation was given by:
\begin{eqnarray*}
P_0 (x_i) &=& C_{ne}(\alpha,\beta,F_i,V)  \; x_i^{-2\beta}  \; \exp\left[\frac{2F_i}{V^2}\frac{x_i^{1+\alpha-2\beta}}{1+\alpha-2\beta}-\frac{\phi_0}{V^2}\frac{x_i^{2-2\beta}}{1-\beta}\right] \quad \textrm{if} \quad  2\beta \ne 1+\alpha, \\
P_0 (x_i) &=& C_e(\alpha,\beta,F_i,V)  \; x_i^{\frac{2F_i}{V^2} -2\beta}  \; \exp\left[\frac{\phi_0}{V^2}\frac{x_i^{2-2\beta}}{1-\beta}\right] \quad \textrm{if} \quad  2\beta = 1+\alpha,
\end{eqnarray*}
where $\phi_0 = \left(\sum_i F_i^{\frac{1}{1-\alpha}}\right)^{1-\alpha}$ and $C_{ne}$ together with $C_{e}$ were integrals solved numerically for the parameters of interest. The ordered phase occurred when the solution had a maximum in the physical interval $(0<x_i<1)$. For a larger V, the transition to a disordered phase happened when the maximum shifted to the unphysical region $x_i<0$, which sets the phase transition region $V(\alpha,\beta,F_i)$. The phase transition region was calculated analytically in specific cases:
\begin{eqnarray*}
F_i^2 &=& 4 \beta \phi_0 V^2\quad \textrm{if} \quad  \beta = \alpha \neq 1,\\
F_i &=& \beta V^2\quad \textrm{if} \quad  2\beta = 1+\alpha,
\notag
\end{eqnarray*}
where the first case, simplified to $F = 3 V^2$ if $\beta = 0.75$ and the fitness of this taxon dominated in $\phi_0$. 
In many physical systems (Brownian motion is the classic example \cite{Einstein}), the two terms of the Langevin's equation are related.  The \emph{Fluctuation Dissipation Theorem} states a general relationship between the response to an external disturbance and the internal fluctuations of the system\cite{FD}. The theorem can be used as the basic formula to derive the fitness from the analysis of fluctuations in the microbiota, assuming that it is in equilibrium (the ordered phase).

%%
\subsection*{Standardization} \label{sec:stan}
In order to properly show all the studies under common axes, we decided to standardize the Taylor parameters using the group of healthy subjects for every single study independently. With this approach, all the studies can be visualized in a shared plot with units of Taylor--parameter standard--deviation on their axes.

For a Taylor's parameter, e.g. $V$, the estimate of the mean ($\widehat{V}$) for the healthy subpopulation, composed of $h$ subjects, is:
\begin{linenomath}
$$\widehat{V} = \frac{1}{W_1}\sum_{i=1}^h V_i \omega_i=\sum_{i=1}^h V_i \omega_i$$
\end{linenomath}
as $W_1=\sum_i^h \omega_i=1$, since $\omega_i$ are normalized weights calculated as:
\begin{linenomath}
$$\omega_i = \frac{\frac{1}{\sigma^2_{V_i}}}{\sum_i^h\frac{1}{\sigma^2_{V_i}}}$$
\end{linenomath}
$\sigma_{V_i}$ being the estimation of the uncertainty in $V_i$ obtained together with $V_i$ from the X-weighted power-law fit described in Section \ref{sec:X-w}, for healthy subjects.

@@

%%
\subsubsection*{Sample selection}
We chose studies about relevant pathologies, containing metagenomic sequencing time data series of bacterial populations from humans in various healthy and non-healthy states. Only those subjects who had three or more time points of data available in databases were selected. The study by Caporaso \textit{et al.} study \cite{moving} was selected as it featured two healthy subjects measured over a very long time-span, with almost daily sampling. The study of Faith \textit{et al.} \cite{LEA} was selected given the BMI differences between subjects. Moreover, some of them followed diets which could be treated as system perturbations. Only those subjects whose BMI was normal or overweight were considered healthy. The study by Smith \textit{et al.} \cite{kwashiorkor} was selected for both the age of the patients and the rare disease. Regarding kwashiorkor, we considered only the discordant twins, and deemed subjects unaffected by kwashiorkor as being healthy in each pair of patients. The study by David \textit{et al.} \cite{diet} was selected for its differential diets. The healthy period  was considered to be the initial samples of each subject before starting the diet, while the remaining time points were considered as perturbations. Dethlefsen and Relman’s work \cite{antibiotic} was selected due to the interesting treatment of two antibiotic intakes of the same antibiotic by three different subjects. The healthy period was considered to correspond only to those times before any antibiotic treatment, whereas the periods during and after antibiotic intake were considered as perturbations. The work by David \textit{et al.} \cite{hostlife} was selected due to the comprehensive longitudinal data that it provides, plus its complete metadata and the interesting events experienced by both subjects (an infection and a trip abroad). The healthy period was taken from time points before or after each event. Finally, we also considered a study from our group carried out by Durban \textit{et al.} \cite{IBS} in which the healthy subjects were considered as those who did not suffer from irritable bowel syndrome, while the patients who had this disease were taken as perturbations.

@@

%%
\subsubsection*{16rRNA sequences processing}
Reads from the selected studies were first quality filtered using the FastX toolkit\cite{FASTX}, allowing only those reads scoring over 25 for quality in  75\% of the complete sequence. 16S rRNA reads were then clustered at a 97\% nucleotide sequence identity (97\% ID) into operational taxonomic units (OTUs), using the QIIME software package\cite{QIIME} (version 1.8). We followed an open reference OTU picking workflow in all cases. The clustering method used was UCLUST, and the OTUs were matched against the Silva database\cite{SILVA} (version 111, July 2012) and were assigned to a taxonomy with a UCLUST-based consensus taxonomy assigner. The parameters used in this step were: similarity 0.97, prefilter percent id 0.6, maximum accepts 20, maximum rejects 500.

%%
\subsubsection*{Metagenomic sequences processing}
Shotgun metagenomic sequences were analyzed with LMAT (Livermore Metagenomics Analysis Toolkit) software package\cite{LMAT} (version 1.2.4, with Feb'15 release of the LMAT-Grand database). LMAT was run using a Bull shared-memory node belonging to the team's HPC (high performance computing) cluster. It was equipped with 32 cores (64 threads available using Intel Hyper-Threading Technology) as it has two Haswell-based Xeons (22 nm technology), the E5-2698v3@2.3 GHz, sharing half a tebibyte of DRAM memory. This node is also provided with a PCIe SSD card as NVRAM, the Micron P420m HHHL, with 1.4 TB, and 750000 reading IOPS, 4 KB, achieving 3.3 GB/s. The computing node was supplied with a RAID-0 (striping) scratch disk area. We used the ``Grand'' database\cite{LMAT2}, released in Feb'15, provided by the LMAT team, where ``Grand'' refers to a huge database that contains k-mers from all the viral, prokaryote, fungal and protist genomes present in the NCBI database, plus the Human reference genome (hg19), plus GenBank Human, plus the 1000 Human Genomes Project (HGP) (this represents about 31.75 billion k-mers occupying 457.62 GB)\cite{LMAT2}. Before any calculations were made, the entire database was loaded into the NVRAM. With this configuration, the observed LMAT sustained sequence classification rate was $20\unclassrate$. Finally, it is worth mentioning that a complete set of Python scripts  was developed as back-end and front-end of the LMAT pipeline in order to manage the added complexity of time series analysis (\href{https://github.com/DLSteam/MAUS_scripts}{[https://github.com/DLSteam/MAUS\_scripts]}).

@@

When fitting the power-law of std vs. mean, we took into account that every mean has uncertainty and can be estimated for a sample size $n$ by the SEM (\emph{Standard Error of the Mean}). Here, the uncertainties affected the independent variable, so the fit was not as trivial as a Y-weighted fit, where the uncertainties affect the dependent variable. A standard approach to perform this fit is: a) to invert the variables before applying the weights, b) then perform the weighted fit and, finally, c) revert the inversion. This method is deterministic, but the approximate solution worsens with smaller coefficients of determination. To overcome this limitation, we developed a stochastic method using a bootstrapping-like strategy that avoided inversion and was applicable regardless of the coefficient of determination.

### 2017-02-21 mSystems/rev2/mSys_discussion_rev2.tex
mSys: several figs&text changes which lead to REV2 DRAFT delta
On branch master changes to be committed:
	modified:   Response to Reviewer Comments.docx
	modified:   figs/Fig1.eps
	modified:   figs/Fig4.eps
	modified:   figs/Fig5.eps
	new file:   figs/HostLifeStyle_StoolA_day72to122_abroad_TAILLESS_Rank.pdf
	new file:   figs/HostLifeStyle_h_StoolA_day0to70_before_TAILLESS_Rank.pdf
	new file:   figs/HostLifeStyle_h_StoolA_day123to256_returned_TAILLESS_Rank.pdf
	new file:   figs/HostLifeStyle_h_StoolA_day257to364_after_TAILLESS_Rank.pdf
	modified:   figs/_ed_Fig1.pdf
	modified:   figs/_ed_supfig

One of the mains features of this work is to have demonstrated that, independently of its condition, microbiota follows Taylor's law. We have seen that the value of the scaling index in each case is always less than the unity (using standard deviation as the measurement for dispersion), which provides us with information about the community structure. This means that, in relative terms, the most abundant elements in the population are less volatile to perturbations than the less abundant ones. The explanation for this universal pattern is not clear although some hypotheses have been tested in other studies, such as the presence of negative interactions in the population \cite{kilpatrick}, and a demonstration that this may depend on reproductive correlation \cite{ballantyne}. Nevertheless, none of these explanations are sufficient when we are talking about microbiota, as the reproduction term is diffuse, the interactions between its components are not only based on competition \cite{joao, mehta, bucci}, and that even that kind of negative interaction may not effectively yield values less than the unity when referring to a bacterial species \cite{ramslayer}. Anyhow, the values obtained in all cases were very similar from one to another, which could suggest that the community structure is preserved throughout the different scenarios we studied.

The second parameter provides information about noise and can be directly linked to the variability or fluctuation amplitude of the population over time. It is a direct estimator of the stability of the system under study. As we have shown above, the healthy subset of each study has lower variability than the non--healthy subset, when dealing with adult individuals. Interestingly, the variability parameter was higher in the healthy subset in the study of the discordant twins suffering from kwashiorkor disease \cite{kwashiorkor}. In this regard, it has been shown that infant microbiota needs to develop toward a definite, adult state \cite{koenig}. This implies that temporal variability is greater in children compared to a healthy adult state, which should be temporally stable. Thus, our results could point to the need for this variability in order to reach that adult state. Furthermore, as we wanted to see how this variability behaved over time, we calculated the evolution of this parameter for the samples which had enough time sampling. As shown in Figure \ref{fig:tempevo1}, the variability of microbiota fluctuated over time. It is interesting to note in Figure \ref{fig:tempevo2} how this parameter captured the two antibiotic intakes in one of the patients from the study by Dethlefsen and Relman \cite{antibiotic}, especially in that there seems to be some kind of a resilience process in the microbiota due to the lower variability increase in the second antibiotic intake.

The primary hypothesis of this work is that, in adults, having a healthy microbiota means that the population is stable over time and does not move into a state where it is highly susceptible to external or internal perturbations, causing a dysbiotic state in the microbiota. In order to use the valuable information provided by the empirical law of Taylor's work, we proposed the use of Langevin's equation to model how ranking stability evolved over time. While the system noise component can be directly measured as its variability, the other main term needs to be inferred from the model. This term, which we named "fitness", is the one that enables the system to be stable in the face of potential perturbations. In ecological terms, this could represent the nature of interactions that are present among bacteria, between bacteria and other minority populations, such as fungi or archaea, between bacteria and the viral component in microbiota, and interactions between the host and the whole microbiota. As this is a first step to model the temporal stability of microbiota, and given its complex nature, we calculated fitness using the Fluctuation Dissipation Theorem as a first approximation\cite{FD}. Thus, the fitness of microbiota will still need to be modeled in future works in order to make the model more accurate and give it a higher predictive power.

By solving Langevin's differential equation, we can obtain a phase diagram where each microbiota sample can be placed according to its fitness and variability into one of two phases, according to the ranking stability of the system. As we can see in the phase--space in Figure \ref{fig:main3}, three different conditions that could occur are shown. First, we could have a healthy microbiota with some fluctuations, as shown by one of the subjects of Caporaso \emph{et al.}'s study \cite{moving}. Because this case would have good fitness, its temporal variability would not place the microbiota in the unstable phase of the diagram. Second, we have a subject from the study by Dethlefsen and Relman \cite{antibiotic} who was perturbed twice by an antibiotic intake. His microbiota underwent sufficient change so as to lose its stability, and hence be placed in the unstable part. In this location, it is more sensitive to potential perturbations such as, for example, opportunist infections. In the third and last condition, the subject was already in the unstable phase due to a health issue, i.e. IBS. This can be observed in one of the patients from Durban \emph{et al.}'s study \cite{IBS}. In addition, it was shown that this subject's health status improved during the time the experiment was carried out, implying that his microbiota also recovered the stability it had lost. It is interesting to note that in the subject's health from the study made by David \emph{et al.} \cite{hostlife} who suffered a Salmonella infection during the experiment, there was a significant shift in variability and a final recovery from the perturbed state (see Supplementary Figure S\ref{supfig:HLS_xWSummary}).

@@

Specifically, the analysis of the rank stability of different periods of time belonging to the individual \emph{A} in the host lifestyle study \cite{hostlife}, suggests that the presence of \emph{rank stability islands} among medium-ranked taxa is an interesting feature. Interestingly, this stability was compromised when the period was not an ordinary one, suggesting that those taxa were sensitive to changes in the lifestyle. Among the genera identified as \emph{rank stability islands}, \emph{Lachnobacterium} and \emph{Clostridium} were catalogued as genera predictive of dysbiosis in the work of Larsen and Dai \cite{rsi_dysbiosis}, which analysed the same dataset \cite{hostlife}. Furthermore, very recently, it has been confirmed a clear relationship between \emph{Actinomyces} and conventional adenoma \cite{rsi_actino}, one of the two main precursors of the colorectal cancer. Finally,\emph{Eggerthella} is an opportunistic pathogen that is often associated with serious gastrointestinal pathology \cite{rsi_egg}.

It could be brought into question the role of these taxa as key players in the phase transition of the microbiota, or whether they are more susceptible to perturbations than the most abundant. The types of interactions that could sustain this particular behavior are not clear, as these non-abundant taxa are not usually included in dynamic studies in order to obtain a community matrix. Further experiments and data analysis are needed to clarify whether \emph{rank stability islands} are a widespread feature of microbiotas and whether they appear at lower taxonomic levels too.

However, we have to be aware that the hypothesis above is too simplistic to be directly related to reality. It has been demonstrated that the situation is more complex than the outlook provided which separate healthy people from non-healthy people just by compositional terms, as Moya and Ferrer underlined in their recent review \cite{Moya_trends}. There are several different feasible scenarios in which we can consider microbiota as being stablem irrespective of their compositional evolution over time. For example, depending on its ability to recover its initial composition (resilience), or whether it can recover its original function despite its composition (functional redundancy). What we have shown in this work could be explained as the transition of stable microbiota into a state of dysbiosis.

As a first step towards understanding microbiota stability, the model presents some limitations and there is still work to do. From a biological perspective, many questions arise from this work. We have observed the same pattern in Taylor's parameters in all the different conditions we studied, but a pertinent question is whether this is really a universal feature in the huge diversity of microbial niches. Furthermore, another relevant question centers on which mechanisms are involved in maintaining the population structure. The nature of the interactions between the elements of community is surely of great importance in this matter, and this is related to the fitness of the community, as mentioned above. How we should address community fitness is not clear, but works such as the one by Tikhonov \cite{tikhonov} could point us in the right direction to unravel the complexity of microbiota.

>>>>

% !TeX root = ./mSys_MAIN_rev2.tex
One of the highlights of this study is that is shows, independently of its condition, that microbiota follows Taylor’s law. We have seen that in each case the value of the scaling index is always less than the unity (using standard deviation as the dispersion measurement), which provides us with information about the community structure. This means that, in relative terms, the most abundant elements in the population are less volatile to perturbations than the less abundant ones. The explanation for this universal pattern is not clear although some hypotheses have been tested in other studies, such as the presence of negative interactions in the population \cite{kilpatrick}, and a demonstration that this may depend on reproductive correlation \cite{ballantyne}. Nevertheless, none of these explanations are sufficient when it comes to microbiota, as the term reproduction is diffuse the interactions between its components are not only based on competition \cite{joao, mehta, bucci}. Moreover, even such negative interaction may not effectively yield values less than the unity when referring to a bacterial species (41). Nonetheless, the values obtained in all cases were very similar to each other, which may suggest that the community structure is preserved throughout the different scenarios studied herein.

The second parameter provides information about noise and can be directly linked to the variability or fluctuation amplitude of the population over time. It is a direct estimator of the stability of the system under study. As we have shown above, the healthy subset of each study has lower variability than the non-healthy subset, when dealing with adult subjects. Interestingly, the variability parameter was higher in the healthy subset in the study of discordant twins suffering from kwashiorkor disease \cite{kwashiorkor}. In this respect, research has shown that infant microbiota needs to develop toward a definite, adult state \cite{koenig}. This implies that temporal variability is greater in children than in a healthy adult, which should be temporally stable. Thus, our results could indicate this variability is necessary in order to reach that adult state. Furthermore, as we wanted to see how this variability shifted over time, we calculated the changes in this parameter for the samples which had enough time sampling. As shown in Figure \ref{fig:tempevo1}, the variability of microbiota fluctuated over time. Interestingly, Figure \ref{fig:tempevo2} shows how this parameter reflected the two antibiotic intakes in one of the patients in the study by Dethlefsen and Relman \cite{antibiotic} particularly the apparent resilience of the microbiota due to the reduced increase in variability during the second antibiotic intake.

The primary hypothesis of this work is that, in adults, having a healthy microbiota means that the microbial population is stable over time.  This stability means the microbiota does not shift and become susceptible to external or internal perturbations causing dysbiosis. In order to use the valuable information provided by the empirical law of Taylor’s work, herein we have proposed the use of Langevin’s equation to model how stability ranking changed  over time. While the system noise component can be directly measured as its variability, the other main term needs to be inferred from the model. This term, which we have named "fitness", enables the system to remain stable when confronted with potential perturbations. In ecological terms, this could represent the nature of interactions present among bacteria, between bacteria and other minority populations, such as fungi or archaea, between bacteria and the viral component in microbiota, and interactions between the host and the whole microbiota. As this is a first step to model the temporal stability of microbiota, and given its complex nature, we calculated fitness using the Fluctuation Dissipation Theorem as a first approximation \cite{FD}. Thus, future works are required to model the fitness of microbiota in order to provide a more accurate model with higher predictive power.

By solving Langevin’s differential equation, we obtain a phase diagram where each microbiota sample can be placed by its fitness and variability into one of two phases, according to the stability ranking of the system. As shown by the phase--space in Figure \ref{fig:main3}, three different conditions can occur. The first is a healthy microbiota with some fluctuations, as shown by one of the subjects in Caporaso \emph{et al.}'s study \cite{moving}. Because this case would have good fitness, its temporal variability would not place the microbiota in the unstable phase of the diagram. Secondly, we have a subject from the study by Dethlefsen and Relman \cite{antibiotic} whose microbiota was perturbed twice by an antibiotic intake, undergoing sufficient change so as to lose its stability, and hence be placed in the unstable part. In this location, it is more sensitive to potential perturbations such as, for example, opportunistic infections. In the third and last condition, the subject was already in the unstable phase due to a health issue, i.e. IBS. This can be observed in one of the patients in Durban \emph{et al.} \cite{IBS}. In addition, it was shown that this subject’s health status improved during the experiment, implying that his/her microbiota also recovered stability. Interestingly, in the study made by David \emph{et al.} \cite{hostlife} the subject who had a Salmonella infection during the experiment underwent a significant shift in variability with eventual recovery from the perturbed state (see Supplementary Figure S\ref{supfig:HLS_xWSummary}).

@@

Specifically, in the host lifestyle study \cite{hostlife}, the presence of \emph{rank stability islands} among medium--ranked taxa is an interesting feature revealed by the analysis of rank stability at different time periods in subject \emph{A}. Interestingly, this stability was compromised when the period was not an ordinary one, suggesting that those taxa were sensitive to changes in lifestyle. Among the genera identified as \emph{rank stability islands}, \emph{Lachnobacterium} and \emph{Clostridium} were catalogued as genera predictive of dysbiosis in the work of Larsen and Dai \cite{rsi_dysbiosis}, which analysed the same dataset \cite{hostlife}. Furthermore, research has recently confirmed a clear relationship between \emph{Actinomyces} and conventional adenoma \cite{rsi_actino}, one of the two main precursors of the colorectal cancer. Finally, \emph{Eggerthella} is an opportunistic pathogen that is often associated with serious gastrointestinal pathology \cite{rsi_egg}.

One might  question the role of these taxa as key players in the phase transition of the microbiota and wonder whether they are more susceptible to perturbations than the most abundant taxa. The types of interactions that could sustain this particular behavior are unclear, as these non-abundant taxa are usually excluded from dynamic studies to obtain a community matrix. Further experiments and data analysis are needed to clarify whether \emph{rank stability islands} are a widespread feature of microbiotas and whether they appear at lower taxonomic levels too.

Notwithstanding the above, we should be aware that the above hypothesis is too simplistic to directly apply to reality. Indeed, the situation is more complex than the idea that healthy people can be distinguished from non-healthy people in solely compositional terms, as highlighted by Moya and Ferrer in their recent review \cite{Moya_trends}. There are several feasible scenarios in which we can consider microbiota to be stable, irrespective of its compositional shifts over time. For example, it may depend on the ability of the microbiota to recover its initial composition (resilience), or its ability to recover its original function despite its composition (functional redundancy). What we have shown in this work could be explained as the transition of stable microbiota into a state of dysbiosis.

This is a first step towards understanding microbiota stability, although the model presents some limitations and thus further research is required. From a biological perspective, many questions arise from this work. We have observed the same pattern in Taylor’s parameters under all the conditions studied, but a pertinent question is whether this is really a universal feature in the hugely diverse microbial niches. Furthermore, another relevant question relates to mechanisms involved in maintaining the population structure.  Undoubtedly, the nature of the interactions between community components has a great bearing on this issue, and this is related to community fitness, as mentioned above. How we should address community fitness remains unclear, but studies like the one by Tikhonov  \cite{tikhonov} could point us in the right direction and help us to unravel the complexity of microbiota and its relationship to host health.
